Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-15 paper_io_sim.tex
Edited abstract

    % Henri Rewrote quite  a bit in the next paragraph here to spell out a bit
    % more why we need simulation
    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data storage management techniques.  While
    infrastructures with sufficient compute power are available to
    process large datasets, the I/O bottleneck remains.  A
    well-known approach for reducing the overhead of
    I/O operations is to use the Linux page cache.  It is thus crucial
    to study experimentally the impact of the Linux page cache on the
    performance of Big Data applications. Given the difficulties with and shortcomings of
    real-world experiments, a popular approach for application
    performance studies is to use simulation.  While many simulation
    frameworks have been developed for this purpose, they do not 
    simulate page caching fully, or even at all.  As a result, using
    these frameworks for studying the performance of data-intensive
    applications leads to inaccurate simulation results.

    In this paper, we propose a I/O simulation model that characterizes
    the key features of the page cache. We have implemented this model
    as  part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework.  Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os.  We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        The Linux Page Cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.

>>>>

    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data processing solutions. While infrastructures
    with sufficient compute power are available,
    the I/O bottleneck remains. The Linux page cache is an efficient
    approach to reduce I/O overheads, but few
    experimental studies of its interactions with Big Data applications exist,
    partly due to limitations associated with
    real-world experiments. Simulation is a popular approach to address
    these issues, however, existing simulation frameworks do not simulate
    page caching fully, or even at all.  As a result, simulation-based
    performance studies of data-intensive applications leads to inaccurate
    results.

    In this paper, we propose an I/O simulation model that includes
    the key features of the page cache. We have implemented this model
    as part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework. Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os. We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        The Linux page cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.
